Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-12 MS/ms.Rmd
Updated PRISMA text
Fixed up Patrice's comments

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. Then, we show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. We formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate nuisance heterogeneity at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

Nuisance heterogeneity can get in the way of understanding real biological patterns that interest comparative physiologists. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest then this variable can be formally incorporated in statistical models. If, however, it is not the primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above. Nonetheless, there is limited discussion about the ways in which it can be dealt with in the literature.

@@

We searched the Scopus database between May 16 and 18, 2021, with a follow up search on September 1, 2021 (see the Supplementary Information). After a series of pilot searches (detailed in Supplementary Materials) to refine our search string [See @Foo2021 for an overview of this process], we searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

All title, abstract and full text screening was done by two people (DWAN and ML). Any disagreements were resolved by discussion to reach a consensus. Out of the 417 papers originally identified, 79 full texts were screened for eligibility, and we included a total of 62 for final data extraction. For full details on the search strings used, search dates, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the Supplementary Information.

@@

# Maybe add some examples for each effect size in a new column. SHould help
eg <- c("CTmin, CTmax, LC50, LT50, Metabolic Rate (MR)",
        "Variability in CTmin, CTmax, Metabolic Rate (MR)",
        "Ratio between pollutant exposed treatment (e.g., BPA-exposed group) and control (no pollutant)",
        "Difference in immune response between males and females or performance difference in the presence of stressor compared to absence of stressor",
        "Relationship between sex hormones and immune responses or metabolic rate and behaviour")

@@

Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions? Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much T has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

>>>>

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. Then, we show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. We formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate nuisance heterogeneity at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that expanding the meta-analytic toolkit will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore, be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

Nuisance heterogeneity can get in the way of understanding real biological patterns that interest comparative physiologists. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not the primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above. Nonetheless, there is limited discussion about the ways in which it can be dealt with in the literature.

@@

We searched the Scopus database between May 16 and 18, 2021, with a follow up search on September 1, 2021 (see the Supplementary Information). After a series of pilot searches (detailed in Supplementary Materials) to refine our search string [See @Foo2021 for an overview of this process], we searched for studies that included: 'meta-analysis', 'meta regression', 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

All title, abstract and full text screening was done by two people (DWAN and ML). Any disagreements were resolved by discussion to reach a consensus. Out of the 426 papers originally identified, 80 full texts were screened for eligibility, and we included a total of 63 for final data extraction. For full details on the search strings used, search dates, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the Supplementary Information.

@@

# Maybe add some examples for each effect size in a new column. SHould help
eg <- c("CTmin, CTmax, LC50, LT50, Metabolic Rate (MR)",
        "Variability in CTmin, CTmax, Metabolic Rate (MR)",
        "Ratio between pollutant exposed treatment (e.g., BPA-exposed group) and control (no pollutant)",
        "Difference in immune response between males and females, performance difference in the presence of stressor compared to absence of stressor",
        "Relationship between sex hormones and immune responses or metabolic rate and behaviour")

@@

Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions? Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much testosterone (T) has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.
